Constructing a universe for the setoid model

### 2020-07-06 paper.tex
small fixes

In Section~\ref{setoid-universe} we discuss the setoid model and various design
choices related to it. We then recall inductive-recursive universes, and the way
they can be equivalently defined as a plain inductive definition, in
Section~\ref{ir-universes}.
%
We then provide, in Section~\ref{ir-setoid-universe}, a first complete
definition of the setoid universe using a special form of induction-recursion.
This form of induction-recursion is not known to be reducible to plain inductive types.
Then we describe an alternative definition of the universe in
Section~\ref{ii-universe}, that does not rely on induction-recursion but instead
on infinitary induction-induction. This inductive-inductive encoding of the
universe is obtained from the inductive-recursive one, following the method of
Section~\ref{ir-universes}.
%
We end the series of universe constructions with
Section~\ref{inductive-universe}, where we outline a purely inductive definition
of the setoid universe, obtained from the inductive-inductive one.

@@

In this section we give a first definition of the setoid universe, as a direct
generalization of the simple inductive-recursive definition just shown. We only
consider a very small universe with bool type $\mathbbm{2}$ and $\Pi$ for
simplicity; a more realistic universe can be found in the Agda formalization,
and includes more type formers.

To construct the universe of setoids in the setoid model, we first of all need
to define a type $\Uty : \Ty\ \GG$ for every $\GG : \Con$, and for every $A :
\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling
Section~\ref{setoid-model}, these are essentially record types made of several
components. Since $\Uty$ is a closed type, it requires the same data of a
setoid; in particular, we need a type of codes together with an equivalence
relation reflecting equality of codes, in addition to proofs that these are
indeed equivalence relations:
%

@@

We will follow the same method outlined in Section~\ref{ir-universes}. In
addition to $\ad{in-U}$ for defining $\U$, we also introduce a family
$\ad{in-U∼}$ of binary relations between types in the universe, from which we
then define $\_\sim_{\setoidU}\_$.
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \mType \to \mType_1 \\
  & \textsf{data}\ \ad{in-U∼} : \{A\ A' : \mType\} \to \ad{in-U}\ A \to \ad{in-U}\ A' \to (A \to A' \to \mProp) \to \mType_1
\end{align*}

@@

It is known that finitary inductive-inductive definitions can be reduced to
inductive families \cite{iit-erasure,iit-to-ix,induction-is-enough}.
%
To illustrate the idea, let us consider a well-known example of finitary
inductive-inductive type, the intrinsic encoding of type theory in type theory
itself. Actually, we only need to consider the type of contexts $\textsf{Con} :
\mType$ and the type of types $\textsf{Ty} : \textsf{Con} \to \mType$; since the
latter is indexed over the former, this is already an example of
induction-induction.

@@

The general method to eliminate induction-induction is to split the original
inductive-inductive types into a type of codes and an associated well-formedness
predicate. In our \textsf{Con/Ty} example, these would be respectively given by
codes $\textsf{Con}_0, \textsf{Ty}_0 : \mType$ and predicates $\textsf{Con}_1 :
\textsf{Con}_0 \to \mType, \textsf{Ty}_1 : \textsf{Con}_0 \to \textsf{Ty}_0 \to
\mType$.

@@

The next step is to prove that these relations are functional, by induction on
the untyped codes $\textsf{Con}_0$ and $\textsf{Ty}_0$. From this result,
defining the eliminators is immediate.

@@

The obvious challenge in successfully completing this reduction is to avoid the
need for extensionality in the metatheory. In fact, consider the simple
infinitary inductive-inductive type obtained from the previous $\textsf{Con/Ty}$
example by replacing the finitary constructor $\Pi$ with an infinitary one: $\Pi
: \{\GG : \textsf{Con}\} \to (\mathds{N} \to \textsf{Ty}\ \GG) \to
\textsf{Ty}\ \GG$. Already with this simple example, we run into problems as
soon as we try to define the eliminator. One issue is that the definition of the
eliminator relies of a proof that the well-typing predicates $\ad{in-U}_1,
\ad{in-U∼}_1$ are propositional, that is, any two of their elements are equal.
%
Without further assumptions this proof can only be done by induction, and
requires function extensionality since these predicates include higher-order
constructors.

@@

Using $\mProp$ avoid the issue of proving propositionality altogether, since the
predicates are now propositional by definition.
%
However, it introduces a different issue: $\ad{in-U}_1$ and $\ad{in-U∼}_1$ give
rise to equational constraints on their indices, in the form of proofs of the
$\mProp$-valued identity type. The definition of the eliminators for $\ad{in-U}$
and $\ad{in-U∼}$ relies on the ability to transport along these proofs, hence
the need to extend our metatheory with a primitive, strong form of transport for
$\textsf{Id}$.\footnote{Note that this issue cannot be solved by expressing the
  equational constraints with an identity type in $\mType$, since the
  well-typing predicates force it to necessarily be in $\mProp$.}

@@

Another question regards the relationship between SeTT \cite{mpc19} and XTT
\cite{xtt}. Both systems are syntactic representations of the setoid model with
similar design choices, like definitional proof-irrelevance. We would like to
know whether their respective notions of models are equivalent, that is, whether
we can obtain an XTT model from a SeTT model, and vice versa. Since XTT
universes support universe induction, it is possible that we would need to
extend our own universe with the same principle (see discussion in
Section~\ref{setoid-model} and the previous paragraph). A further question is
whether this mapping of models is functorial.

>>>>

In Section~\ref{setoid-universe} we discuss the setoid model and various design
choices related to it. We then recall inductive-recursive universes, and the way
they can be equivalently defined as a plain inductive definition, in
Section~\ref{ir-universes}.
%
We then provide, in Section~\ref{ir-setoid-universe}, a first complete
definition of the setoid universe using a special form of induction-recursion.
This form of induction-recursion is not known to be reducible to plain inductive types.
Then we describe an alternative definition of the universe in
Section~\ref{ii-universe}, that does not rely on induction-recursion but instead
on infinitary induction-induction. This inductive-inductive encoding of the
universe is obtained from the inductive-recursive one, inspired by the method of
Section~\ref{ir-universes}.
%
We end the series of universe constructions with
Section~\ref{inductive-universe}, where we outline a purely inductive definition
of the setoid universe, obtained from the inductive-inductive one.

@@

In this section we give a first definition of the setoid universe, as a direct
generalization of the simple inductive-recursive definition just shown. We only
consider a very small universe with bool type $\mathbbm{2}$ and $\Pi$ for
simplicity; a more realistic universe that includes more type formers
can be found in the Agda formalization.

To construct the universe of setoids in the setoid model, we first of all need
to define a type $\Uty : \Ty\ \GG$ for every $\GG : \Con$, and for every $A :
\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling
Section~\ref{sec:setoid-model}, these are essentially record types made of several
components. Since $\Uty$ is a closed type, it requires the same data of a
setoid; in particular, we need a type of codes together with an equivalence
relation reflecting equality of codes, in addition to proofs that these are
indeed equivalence relations:
%

@@

We will follow the method outlined in Section~\ref{ir-universes}. In
addition to $\ad{in-U}$ for defining $\U$, we also introduce a family
$\ad{in-U∼}$ of binary relations between types in the universe, from which we
then define $\_\sim_{\setoidU}\_$.
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \mType \to \mType_1 \\
  & \textsf{data}\ \ad{in-U∼} : \{A\ A' : \mType\} \to \ad{in-U}\ A \to \ad{in-U}\ A' \to (A \to A' \to \mProp) \to \mType_1
\end{align*}

@@

It is known that finitary inductive-inductive definitions can be reduced to
inductive families \cite{iit-erasure,iit-to-ix,induction-is-enough}.
%
To illustrate the idea, let us consider a well-known example of a finitary
inductive-inductive type, the intrinsic encoding of type theory in type theory
itself. Actually, we only consider the type of contexts $\textsf{Con} :
\mType$ and the type of types $\textsf{Ty} : \textsf{Con} \to \mType$; since the
latter is indexed over the former, this is already an example of
induction-induction.

@@

The general method to eliminate induction-induction is to split the original
inductive-inductive types into a type of codes and associated well-formedness
predicates. In our \textsf{Con/Ty} example, these would be respectively given by
codes $\textsf{Con}_0, \textsf{Ty}_0 : \mType$ and predicates $\textsf{Con}_1 :
\textsf{Con}_0 \to \mType, \textsf{Ty}_1 : \textsf{Con}_0 \to \textsf{Ty}_0 \to
\mType$.

@@

The next step is to prove that these relations are functional, by induction on
the untyped codes $\textsf{Con}_0$ and $\textsf{Ty}_0$ \cite{induction-is-enough}. From this result,
defining the eliminators is immediate.

@@

The obvious challenge in successfully completing this reduction is to avoid the
need for extensionality in the metatheory. In fact, consider the simple
infinitary inductive-inductive type obtained from the previous $\textsf{Con/Ty}$
example by replacing the finitary constructor $\Pi$ with an infinitary one: $\Pi
: \{\GG : \textsf{Con}\} \to (\mathds{N} \to \textsf{Ty}\ \GG) \to
\textsf{Ty}\ \GG$. Already with this simple example, we run into problems as
soon as we try to define the eliminator. One issue is that the definition of the
eliminator relies on a proof that the well-typing predicates $\ad{in-U}_1,
\ad{in-U∼}_1$ are propositional, that is, any two of their elements are equal.
%
Without further assumptions this proof can only be done by induction, and
requires function extensionality since these predicates include higher-order
constructors.

@@

Using $\mProp$ avoids the issue of proving propositionality altogether, since the
predicates are now propositional by definition.
%
However, it introduces a different issue: $\ad{in-U}_1$ and $\ad{in-U∼}_1$ give
rise to equational constraints on their indices, in the form of proofs of the
$\mProp$-valued identity type. The definition of the eliminators for $\ad{in-U}$
and $\ad{in-U∼}$ relies on the ability to transport along these proofs, hence
the need to extend our metatheory with a primitive, strong form of transport for
$\textsf{Id}$.\footnote{Note that this issue cannot be solved by expressing the
  equational constraints with an identity type in $\mType$, since the
  well-typing predicates force it to necessarily be in $\mProp$.}

@@

Another question regards the relationship between SeTT \cite{mpc19} and XTT
\cite{xtt}. Both systems are syntactic representations of the setoid model with
similar design choices, like definitional proof-irrelevance. We would like to
know whether their respective notions of models are equivalent, that is, whether
we can obtain an XTT model from a SeTT model, and vice versa. Since XTT
universes support universe induction, it is possible that we would need to
extend our own universe with the same principle (see discussion in
Section~\ref{sec:setoid-model} and the previous paragraph). A further question is
whether this mapping of models is functorial.
